---
title: "Exploring putative H2S producing bacteria in IBD, CRC and healthy individuals"
author: "Domenick J. Braccia"
date: "12/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading Libraries

```{r message=FALSE, warning=FALSE}
library(curatedMetagenomicData)
library(tictoc)
library(dplyr)
library(ggplot2)
library(VennDiagram)
library(UpSetR)
library(patchwork)
library(cowplot)
library(svglite)
library(scales)
library(reshape)
```

## Loading and Processing Data

```{r}
source("process_data.R")
```

## Overlap of cys degrading enzymes

upset plot for each cys degrading gene

```{r, message=FALSE, warning=FALSE}
## NOTE: only way to save this plot is by knitting the doc and clicking saveas on the figure
source("upset.R")
upset(cys_dsr_overlap, 
      sets = colnames(cys_dsr_overlap)[-1], 
      order.by = "freq", decreasing = TRUE,
      matrix.color = c("grey23"), point.size = 2,
      sets.bar.color = c(rep("#dd9e2f", 4), "#1c6bc9", rep("#dd9e2f", 1)),
      main.bar.color = c(rep("#dd9e2f", 3), "#1c6bc9", rep("#dd9e2f", 9)),
      text.scale = 1,
      mainbar.y.label = "Number of hits to genomes in UHGG") 
```

## cys degredation vs. DSR in healthy controls

plotting the relative abundances of bacteria with cysteine degrading and sulfite reducing potential. Abundances come from `curatedMetagenomicData` studies. Here, we are looking at healthy controls in particular.

```{r, message=FALSE, warning=FALSE}
source("healthy_relative_abundances.R")
cys_met_dsr_healthy_RA
```

## In at risk groups (IBD, CRC)

```{r, message=FALSE, warning=FALSE}
source("risk_relative_abundances.R")
(cys_k2_boxplot | dsrAB_k2_boxplot) +
  plot_annotation(title = "curatedMetagneomicData (52 studies)",
                  theme = theme(plot.title = element_text(size = 18)))
(cys_k2_hmp2_boxplot | dsrAB_k2_hmp2_boxplot) +
  plot_annotation(title = "HMP2",
                  theme = theme(plot.title = element_text(size = 18)))
(cys_k2_prism_boxplot | dsrAB_k2_prism_boxplot) +
  plot_annotation(title = "PRISM",
                  theme = theme(plot.title = element_text(size = 18)))
(cys_k2_cib_boxplot | dsrAB_k2_cib_boxplot) +
  plot_annotation(title = "Pediatric Crohn's (Lewis et. al 2015)",
                  theme = theme(plot.title = element_text(size = 22)))
```

## Transcriptomic Analysis

```{r}
source("gene_expression.R")
baseline_boxplot | plant_boxplot | animal_boxplot
```

