---
title: "Prevalence of cysteine degrading bacteria in the healthy human gut"
author: "Domenick J. Braccia"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

TODOs:
* [ ] - Include hmp2 kraken2 RA data into figures 2 & 3
* [X] - (figure 3) subset controls to those associated with the IBD and CRC groups (adenoma is now included)

* [X] - generate prevalence of Sulfate and Sulfite Reducing Bacteria
* [X] - run GutFunFind on known genes and then expand list of cysteine reducing bacteria 
* [X] - import list of cysteine degrading bacteria to this script
* [X] - run same set of analyses on these names as the sulfate / sulfite reducing bac.
* [X] - make boxplots for RA of sulfi(a)te red. bac. and RA of cys. deg. bac.


# Loading Libraries

```{r, message=FALSE, warning=FALSE}
library(curatedMetagenomicData, quietly = TRUE)
library(tictoc)
library(dplyr)
library(ggplot2)
library(VennDiagram)
library(UpSetR)
library(patchwork)
library(cowplot)
library(svglite)
library(scales)
library(reshape)
```


# Loading data

```{r}
if(file.exists("../data/all_data.RDS")) {
  metaphlan2_data <- readRDS("../data/all_data.RDS")
} else {
  tic()
  metaphlan2_data <- curatedMetagenomicData("*metaphlan_bugs_list.stool*", dryrun = FALSE)
  saveRDS(all_data, file = "../data/all_data.RDS")
  toc()
}

## loading kraken2 data from Xiaofang
kraken2_counts <- readRDS("../data/from-xiaofang/kraken2_output.rds")
kraken2_pData <- readRDS("../data/from-xiaofang/metadata.rds")
kraken2_hmp2_counts <- read.csv("../data/from-xiaofang/HMP2-IBD.mpa.txt", sep = "\t", header = TRUE, row.names = 1)
kraken2_hmp2_pData <- read.csv("../data/from-xiaofang/hmp2_metadata.MGX.csv", header = TRUE) # NOTE: `diagnosis` column contains study condition information.
kraken2_hmppilot_counts <- read.csv("../data/from-xiaofang/HMP-IBD-pilot.mpa.txt", sep = "\t", header = TRUE)
kraken2_prism_counts <- read.csv("../data/from-xiaofang/merged.mpa.txt", sep = "\t", header = TRUE)
prism_metadata_combined <- read.csv("../data/prism_combined_metadata_20201209.txt")

## loading prism medatada
# prism_metadata <- read.csv("../data/metadata-prism.txt", sep = "\t", header = TRUE)
# prism_extra_metadata <- read.csv("../data/from-brantley/metadata_PRISM_NLIBD", header = TRUE)
# colnames(prism_extra_metadata)[3] <- "LibraryName"
# prism_combined_metadata <- merge(prism_metadata, prism_extra_metadata, by = "LibraryName")
# write.table(prism_combined_metadata, file = "../data/prism_combined_metadata_20201209.txt", quote = FALSE, sep = "\t", row.names = FALSE)
## all code above was for making the combined metadata file
```

# Exploring data

```{r}
## merging all datasets
metaphlan2_data_merged <- mergeData(metaphlan2_data)
metaphlan2_pData <- pData(metaphlan2_data_merged)
study_cond <- metaphlan2_pData$study_condition
study_cond_NA <- metaphlan2_data_merged[,is.na(metaphlan2_pData$study_condition) ]
```

# Processing and subsetting data

```{r}
# ============================ METAPHLAN2 DATA ============================== #

## removing NAs from pData
metaphlan2_pData <- metaphlan2_pData[!is.na(metaphlan2_pData$study_condition), ]

## subsetting pData to study_condition == "controls"
metaphlan2_pData_controls <- metaphlan2_pData %>% filter(study_condition == "control")

## getting rid of NAs in relative abundance (RA) data
metaphlan2_RA <- exprs(metaphlan2_data_merged)
metaphlan2_RA <- metaphlan2_RA[, !is.na(metaphlan2_pData$study_condition)]

## subsetting RA down to just present in control samples
metaphlan2_RA_controls <- metaphlan2_RA[, colnames(metaphlan2_RA) %in% rownames(metaphlan2_pData_controls)]

# =============================== KRAKEN DATA =============================== #

## computing RA values from count info
kraken2_counts_spp <- kraken2_counts[grep("s__", rownames(kraken2_counts)), ]
kraken2_RA <- prop.table(kraken2_counts_spp, 2) * 100 ## counts -> RA (%)
kraken2_hmp2_counts_spp <-  kraken2_hmp2_counts[grep("s__", rownames(kraken2_hmp2_counts)), ]
kraken2_hmp2_RA <- prop.table(as.matrix(kraken2_hmp2_counts_spp), 2) * 100 ## counts -> RA (%)

## getting rid of RA data whose sample has study_condition == NA
study_cond_NA <- kraken2_pData %>% filter(is.na(study_condition)) %>% rownames()
kraken2_RA <- select(as.data.frame(kraken2_RA), -study_cond_NA)
kraken2_pData <- filter(kraken2_pData, !is.na(study_condition))
kraken2_pData %>% 
  filter(study_condition == "control") -> kraken2_pData_controls
kraken2_RA_controls <- kraken2_RA[, colnames(kraken2_RA) %in% rownames(kraken2_pData_controls)]

## removing duplicate metadata sample entries
kraken2_hmp2_pData <- kraken2_hmp2_pData[-grep("_", kraken2_hmp2_pData$External.ID), ]

## modifying kraken2_hmp2_pData$External.ID column for compatability with RA
# kraken2_hmp2_pData$External.ID.mod <- sapply(strsplit(kraken2_hmp2_pData$External.ID, "_"), "[", 1)
```


# Prevalence of sulf reducing bacteria

## Importing results from GutFunFind

upset plot
```{r}
## comparison of cys / met degrading bac using upsetR
cys_met_feature <- read.table("../results/from-GutFunFind/Cysteine_and_Methionine_Degradation.feature.tsv")
colnames(cys_met_feature) <- c("genome", "func")
cys_met_feature %>% 
  mutate(absent = ifelse(func == "absent", 1, 0),
         methionine = ifelse((func == "methionine" | func == "both"), 1, 0),
         cysteine = ifelse((func == "cysteine" | func == "both"), 1, 0)) %>%
  select(genome, absent, methionine, cysteine) -> cys_met_overlap

# adding diss sulf red bacteria to the mix
dsr_feature <- read.table("../results/from-GutFunFind/Dissimilatory_Sulf_Reduction.feature.tsv")
colnames(dsr_feature) <- c("genome", "func")
dsr_feature %>% 
  mutate(absent = ifelse(func == "absent", 1, 0),
         dsrA = ifelse((func == "dsrA" | func == "dsrAB"), 1, 0),
         dsrB = ifelse((func == "dsrB" | func == "dsrAB"), 1, 0),
         dsrAB = ifelse(func == "dsrAB", 1, 0)) %>%
  select(genome, absent, dsrA, dsrB, dsrAB) -> dsr_overlap

## TODO: combine "absent" columns from both cys met deg and diss sulf red
cm_dsr_overlap <- cbind(cys_met_overlap, dsrAB = dsr_overlap$dsrAB) 
for (i in 1:dim(cm_dsr_overlap)[1]) {
  if (cm_dsr_overlap$dsrAB[i] == 1) {
    # print(cm_dsr_overlap[i, ])
    cm_dsr_overlap$absent[i] <- 0
  }
}

# NOTE: atm, only way to save this image is to knit the document, then "save image as"
upset(cm_dsr_overlap, 
      sets = c("methionine", "cysteine", "dsrAB"), 
      order.by = "freq", decreasing = TRUE, 
      matrix.color = c("grey23"), point.size = 5,
      sets.bar.color = c("blue", "red", "orange"),
      text.scale = 1.4,
      mainbar.y.label = "Number of species with function") 

```

# Prevalence of H2S Producing Bacteria

## Importing results from GutFunFind

Before reporting results similar to the sulfite and sulfate reducing bacteria for cysteine degrading bacteria, I used GutFunFind to query known cysteine degrading enzymes against the 4,644 representative UHGG genomes. 

Initially, I was going through an arduous process of parsing these taxonomic hits and trying to match these hits up to the same format that the metaphlan2 RA taxa names are from `cMD`. Since then, Xiaofang has gone through the trouble of running kraken2 on the `cMD` datasets so that these lineages match up with the relative abundance data. 

**IMPORTANT** I have now separated the processing done to match taxa hits with metaphlan RA data. The processing required to match taxa hits with kraken2 RA data is minimal to non existent, so that is now placed in a separate chunk.

This taxonomic data has been imported from GutFunFind and is located in `results/from-GutFunFind/{function_name}.taxa_hits.txt`.

```{r}
## importing taxa hit information and feature information
dsr_taxa_hits_raw <- read.csv("../results/from-GutFunFind/Dissimilatory_Sulf_Reduction.taxa_hits.txt", sep = "\t", header = FALSE)
cys_met_taxa_hits_raw <- read.csv("../results/from-GutFunFind/Cysteine_and_Methionine_Degradation.taxa_hits.txt", sep = "\t", header = FALSE)
uhgg_genid2taxa <- read.csv("../data/from-xiaofang/spec.txt", sep = '\t', header = FALSE)
uhgg_genid2taxa$V2 <- gsub(";", "|", uhgg_genid2taxa$V2)
```

## Using kraken2 output as relative abundance data

Here, I am starting from raw taxa hit information from GutFunFind and seeing how little parsing I need to do to match up taxa hits with the relative abundance matrix. Above, the taxa hit names required a lot of processing in order to match hit names with relative abundance info from metaphlan, but apparently I will not have to do as much with these RA data.

```{r}
## converting imported df -> vector
dsr_taxa_hits <- as.vector(t(dsr_taxa_hits_raw), mode = "character") 
dsr_taxa_hits <- gsub(";", "|", dsr_taxa_hits)
cys_met_taxa_hits <- as.vector(t(cys_met_taxa_hits_raw), mode = "character")
cys_met_taxa_hits <- gsub(";", "|", cys_met_taxa_hits)

## selecting SRB that contain both dsrA and dsrB
dsr_overlap %>% 
  filter(dsrAB == 1) %>% 
  select(genome) -> dsrAB_genids
dsrAB_taxa <- uhgg_genid2taxa$V2[uhgg_genid2taxa$V1 %in% as.vector(t(dsrAB_genids), mode = "character")]

## subsetting kraken2 RA controls data for SRB
dsr_k2_RA_controls <- kraken2_RA_controls[dsrAB_taxa, ]
dsr_k2_cs_controls <- colSums(dsr_k2_RA_controls)

## subsetting kraken2 RA data for cys_met degrading bacteria
cys_met_overlap %>%
  filter(cysteine == 1) %>%
  select(genome) -> cys_genids
cys_taxa <- uhgg_genid2taxa$V2[uhgg_genid2taxa$V1 %in% as.vector(t(cys_genids), mode = "character")]
cys_k2_RA_controls <- kraken2_RA_controls[cys_taxa, ]
cys_k2_cs_controls <- colSums(cys_k2_RA_controls)

cys_met_overlap %>%
  filter(methionine == 1) %>%
  select(genome) -> met_genids
met_taxa <- uhgg_genid2taxa$V2[uhgg_genid2taxa$V1 %in% as.vector(t(met_genids), mode = "character")]
met_k2_RA_controls <- kraken2_RA_controls[met_taxa, ]
met_k2_cs_controls <- colSums(met_k2_RA_controls)


cys_met_overlap %>%
  filter(cysteine ==1, methionine == 1) %>%
  select(genome) -> cys_met_genids
cys_met_taxa <- uhgg_genid2taxa$V2[uhgg_genid2taxa$V1 %in% as.vector(t(cys_met_genids), mode = "character")]
cys_met_k2_RA_controls <- kraken2_RA_controls[cys_met_taxa, ]
cys_met_k2_cs_controls <- colSums(cys_met_k2_RA_controls)
```



## Using metaphlan2 values as relative abundance data

In order to use the metaphlan2 values as the input to the RA plots, lots of processing needs to be done on the taxa hits returned by GutFunFind. I am keeping all of these processing steps here for plotting later.

cys, met, cys_met
```{r}
cys_met_taxa_hits <- as.vector(t(cys_met_taxa_hits_raw), mode = "character") # df -> vector
cys_met_taxa_hits <- strsplit(cys_met_taxa_hits, ";")

## extracting species names from taxa hits
cys_met_taxa_hits_spp <- vector(mode = "character", length = length(cys_met_taxa_hits)) # create empty caracter vector
for (i in 1:length(cys_met_taxa_hits)) {
  current_spp <- gsub(pattern = "s__", replacement = "", cys_met_taxa_hits[[i]][7])
  cys_met_taxa_hits_spp[i] <- current_spp
}

## subetting feature table to non-absent representative genomes (contained in dsr_taxa_hits_raw)
filter(cys_met_overlap, absent == 0) -> cys_met_overlap_present

## removing empty species names
cys_met_overlap_present <- cys_met_overlap_present[cys_met_taxa_hits_spp != "", ] # simultaneously subsetting the feature table
cys_met_taxa_hits_spp <- cys_met_taxa_hits_spp[cys_met_taxa_hits_spp != ""]
dim(cys_met_overlap_present)
length(cys_met_taxa_hits_spp)

## removing Co-Abundance gene Groups (CAGs), GCAs (?), etc.
if (length(grep("CAG", cys_met_taxa_hits_spp)) == 0) {
  # do nothing
} else if (length(grep("CAG", cys_met_taxa_hits_spp)) > 0) {
  cys_met_overlap_present <- cys_met_overlap_present[-grep("CAG", cys_met_taxa_hits_spp), ]
  cys_met_taxa_hits_spp <- cys_met_taxa_hits_spp[-grep("CAG", cys_met_taxa_hits_spp)]
}
dim(cys_met_overlap_present)
length(cys_met_taxa_hits_spp)
if (length(grep("GCA", cys_met_taxa_hits_spp)) == 0) {
  # do nothing
} else if (length(grep("GCA", cys_met_taxa_hits_spp)) > 0) {
  cys_met_overlap_present <- cys_met_overlap_present[-grep("GCA", cys_met_taxa_hits_spp), ]
  cys_met_taxa_hits_spp <- cys_met_taxa_hits_spp[-grep("GCA", cys_met_taxa_hits_spp)]
}
dim(cys_met_overlap_present)
length(cys_met_taxa_hits_spp)
if (length(grep("sp[0-9]*", cys_met_taxa_hits_spp)) == 0) {
  # do nothing
} else if (length(grep("sp[0-9]*", cys_met_taxa_hits_spp)) > 0) {
  cys_met_overlap_present <- cys_met_overlap_present[-grep("sp[0-9]*", cys_met_taxa_hits_spp), ]
  cys_met_taxa_hits_spp <- cys_met_taxa_hits_spp[-grep("sp[0-9]*", cys_met_taxa_hits_spp)]
}
dim(cys_met_overlap_present)
length(cys_met_taxa_hits_spp)

## subetting based on function hits (complicated for cys met)

### cys degrading
cys_taxa_hits_spp <- cys_met_taxa_hits_spp[cys_met_overlap_present$cysteine == 1]
cys_overlap_present <- cys_met_overlap_present[cys_met_overlap_present$cysteine == 1, ]
dim(cys_overlap_present)
length(cys_taxa_hits_spp)
#### removing oddities from names like "_A", "_X", etc.
for (i in 1:length(cys_taxa_hits_spp)) {
  current_name <- gsub("_[A-Z]{2}", "", cys_taxa_hits_spp[i])
  current_name <- gsub("_[A-Z]_", "_", current_name)
  current_name <- gsub("_[A-Z]", "", current_name)
  # current_name <- gsub("_[A-Z]", "", cys_taxa_hits_spp[i])
  cys_taxa_hits_spp[i] <- current_name
}
dim(cys_overlap_present)
length(cys_taxa_hits_spp)

#### removing "MGYG-HGUT" genomes
cys_taxa_hits_spp <- cys_taxa_hits_spp[-grep("MGYG-HGUT", cys_taxa_hits_spp)]

#### removing duplicates of species names
cys_taxa_hits_spp <- unique(cys_taxa_hits_spp)
print(paste("Number of cys bacteria compatable with metaphlan2 taxa names: ", length(cys_taxa_hits_spp)))

### met degrading
met_taxa_hits_spp <- cys_met_taxa_hits_spp[cys_met_overlap_present$methionine == 1]
met_overlap_present <- cys_met_overlap_present[cys_met_overlap_present$methionine == 1, ]
dim(met_overlap_present)
length(met_taxa_hits_spp)
#### removing oddities from names like "_A", "_X", etc.
for (i in 1:length(met_taxa_hits_spp)) {
  current_name <- gsub("_[A-Z]{2}_", "", met_taxa_hits_spp[i])
  current_name <- gsub("_[A-Z]{2}", "", current_name)
  current_name <- gsub("_[A-Z]_", "_", current_name)
  current_name <- gsub("_[A-Z]", "", current_name)
  # current_name <- gsub("_[A-Z]", "", cys_taxa_hits_spp[i])
  met_taxa_hits_spp[i] <- current_name
}
dim(met_taxa_hits_spp)
length(met_taxa_hits_spp)

#### removing "MGYG-HGUT" genomes
met_taxa_hits_spp <- met_taxa_hits_spp[-grep("MGYG-HGUT", met_taxa_hits_spp)]

#### removing duplicates of species names
met_taxa_hits_spp <- unique(met_taxa_hits_spp)
print(paste("Number of met bacteria compatable with metaphlan2 taxa names: ", length(met_taxa_hits_spp)))

### both degrading
both_taxa_hits_spp <- cys_met_taxa_hits_spp[(cys_met_overlap_present$methionine == 1 & cys_met_overlap_present$cysteine == 1)]
both_overlap_present <- cys_met_overlap_present[(cys_met_overlap_present$methionine == 1 & cys_met_overlap_present$cysteine == 1), ]
dim(both_overlap_present)
length(both_taxa_hits_spp)
#### removing oddities from names like "_A", "_X", etc.
for (i in 1:length(both_taxa_hits_spp)) {
  current_name <- gsub("_[A-Z]{2}_", "", both_taxa_hits_spp[i])
  current_name <- gsub("_[A-Z]{2}", "", current_name)
  current_name <- gsub("_[A-Z]_", "_", current_name)
  current_name <- gsub("_[A-Z]", "", current_name)
  both_taxa_hits_spp[i] <- current_name
}
#### removing "MGYG-HGUT" genomes
both_taxa_hits_spp <- both_taxa_hits_spp[-grep("MGYG-HGUT", both_taxa_hits_spp)]
#### removing duplicates of species names
both_taxa_hits_spp <- unique(both_taxa_hits_spp)
print(paste("Number of cys_met bacteria compatable with metaphlan2 taxa names: ", length(both_taxa_hits_spp)))
```

dsr
```{r}
dsr_taxa_hits <- as.vector(t(dsr_taxa_hits_raw), mode = "character") # df -> vector
dsr_taxa_hits <- strsplit(dsr_taxa_hits, ";")
tmp <- unlist(dsr_taxa_hits)
## extracting species names from taxa hits
dsr_taxa_hits_spp <- vector(mode = "character", length = length(dsr_taxa_hits)) # create empty caracter vector
for (i in 1:length(dsr_taxa_hits)) {
  current_spp <- gsub(pattern = "s__", replacement = "", dsr_taxa_hits[[i]][7])
  dsr_taxa_hits_spp[i] <- current_spp
}

## subetting feature table to non-absent representative genomes (contained in dsr_taxa_hits_raw)
# filter(dsr_overlap, absent == 0) -> dsr_overlap_present

### for use in debugging gutfunfind results for diss. sulf.
# tmp_taxa <- dsr_taxa_hits[dsr_overlap_present$dsrAB == 1] ## NOTE: the version of `dsr_taxa_hits` required for this step
#                                                           ## is the version outputted immediately after the _raw line
# tmp_func <- filter(dsr_overlap_present, dsrAB == 1)
# tmp_prob_genID <- tmp_func$genome[c(2,5,6,7,11,17,18,19,20,26)]
# tmp_problem_taxa <- tmp_taxa[c(2,5,6,7,11,17,18,19,20,26)]
# tmp_accep_genID <- tmp_func$genome[c(1,3,4,8,9,10,12,13,14,15,16,21,22,23,24,25,27)]
# tmp_accepted_taxa <- tmp_taxa[c(1,3,4,8,9,10,12,13,14,15,16,21,22,23,24,25,27)]

## removing empty species names
dsr_overlap_ <- dsr_overlap[dsr_overlap$absent == 0, ]

## removing Co-Abundance gene Groups (CAGs), GCAs (?), etc.
if (length(grep("CAG", dsr_taxa_hits_spp)) == 0) {
  # do nothing
} else if (length(grep("CAG", dsr_taxa_hits_spp)) > 0) {
  dsr_overlap_ <- dsr_overlap_[-grep("CAG", dsr_taxa_hits_spp), ]
  dsr_taxa_hits_spp <- dsr_taxa_hits_spp[-grep("CAG", dsr_taxa_hits_spp)]
}
dim(dsr_overlap_)
length(dsr_taxa_hits_spp)
if (length(grep("GCA", dsr_taxa_hits_spp)) == 0) {
  # do nothing
} else if (length(grep("GCA", dsr_taxa_hits_spp)) > 0) {
  dsr_overlap_ <- dsr_overlap_[-grep("GCA", dsr_taxa_hits_spp), ]
  dsr_taxa_hits_spp <- dsr_taxa_hits_spp[-grep("GCA", dsr_taxa_hits_spp)]
}
dim(dsr_overlap_)
length(dsr_taxa_hits_spp)
if (length(grep("sp[0-9]*", dsr_taxa_hits_spp)) == 0) {
  # do nothing
} else if (length(grep("sp[0-9]*", dsr_taxa_hits_spp)) > 0) {
  dsr_overlap_ <- dsr_overlap_[-grep("sp[0-9]*", dsr_taxa_hits_spp), ]
  dsr_taxa_hits_spp <- dsr_taxa_hits_spp[-grep("sp[0-9]*", dsr_taxa_hits_spp)]
}
dim(dsr_overlap_)
length(dsr_taxa_hits_spp)

## subsetting based on function hits (for example, dsrAB must be present)
`%notin%` <- Negate(`%in%`)
dsr_taxa_hits_spp <- dsr_taxa_hits_spp[dsr_overlap_$dsrAB == 1]
dsr_overlap_ <- dsr_overlap_[dsr_overlap_$dsrAB == 1, ]
dim(dsr_overlap_)
length(dsr_taxa_hits_spp)

## removing genomes without specific species names
dsr_taxa_hits_spp <- dsr_taxa_hits_spp[-grep("MGYG-HGUT", dsr_taxa_hits_spp)]
print(paste("Number of dsr bacteria compatable with metaphlan2 taxa names: ", length(dsr_taxa_hits_spp)))
```

subsetting relative abundances 
```{r}
## CYSTEINE
cys_indicies <- list(mode = "integer")
for (i in 1:length(cys_taxa_hits_spp)) {
  # cys_indicies[[i]] <- grep(cys_taxa_hits_spp[i], rownames(control_RA))
  cys_indicies[[i]] <- grep(cys_taxa_hits_spp[i], rownames(metaphlan2_RA))
}
cys_indicies <- unlist(cys_indicies)
cys_indicies <- unique(cys_indicies) # sanity check, no repeat rows

# subsetting rel. abund. of cysteine deg bacteria
cys_m2_RA_controls <- metaphlan2_RA_controls[cys_indicies, ]
dim(cys_m2_RA_controls)

# removing rows with __t in the taxa name... leading to double counting of RA
cys_m2_RA_controls <- cys_m2_RA_controls[-grep("t__", rownames(cys_m2_RA_controls)), ]
dim(cys_m2_RA_controls)

# computing colSums and calculating XX% prevalance of cys. deg. bac.
cys_m2_cs_controls <- colSums(cys_m2_RA_controls)
cys_presnece_prop <- sum(cys_m2_cs_controls > 0) / length(cys_m2_cs_controls)
print(paste0("Propotion of samples containing at least one met bacteria: ", cys_presnece_prop))

## METHIONINE
met_indicies <- list(mode = "integer")
for (i in 1:length(met_taxa_hits_spp)) {
  met_indicies[[i]] <- grep(met_taxa_hits_spp[i], rownames(metaphlan2_RA))
}
met_indicies <- unlist(met_indicies)
met_indicies <- unique(met_indicies) # sanity check, no repeat rows

# subsetting rel. abund. of cysteine deg bacteria
met_m2_RA_controls <- metaphlan2_RA_controls[met_indicies, ]
dim(met_m2_RA_controls)

# removing rows with __t in the taxa name... leading to double counting of RA
met_m2_RA_controls <- met_m2_RA_controls[-grep("t__", rownames(met_m2_RA_controls)), ]
dim(met_m2_RA_controls)

# computing colSums and calculating XX% prevalance of cys. deg. bac.
met_m2_cs_controls <- colSums(met_m2_RA_controls)
met_presnece_prop <- sum(met_m2_cs_controls > 0) / length(met_m2_cs_controls)
print(paste0("Propotion of samples containing at least one met bacteria: ", met_presnece_prop))

## BOTH (CYSTEINE AND METHIONINE)
both_indicies <- list(mode = "integer")
for (i in 1:length(both_taxa_hits_spp)) {
  both_indicies[[i]] <- grep(both_taxa_hits_spp[i], rownames(metaphlan2_RA))
}
both_indicies <- unlist(both_indicies)
both_indicies <- unique(both_indicies) # sanity check, no repeat rows

# subsetting rel. abund. of cysteine deg bacteria
cys_met_m2_RA_controls <- metaphlan2_RA_controls[both_indicies, ]
dim(cys_met_m2_RA_controls)

# removing rows with __t in the taxa name... leading to double counting of RA
cys_met_m2_RA_controls <- cys_met_m2_RA_controls[-grep("t__", rownames(cys_met_m2_RA_controls)), ]
dim(cys_met_m2_RA_controls)

# computing colSums and calculating XX% prevalance of cys. deg. bac.
cys_met_m2_cs_controls <- colSums(cys_met_m2_RA_controls)
cys_met_presnece_prop <- sum(cys_met_m2_cs_controls > 0) / length(cys_met_m2_cs_controls)
print(paste0("Propotion of samples containing at least one cys_met bacteria: ", cys_met_presnece_prop))

## DSR INDICIES
dsr_indicies <- list(mode = "integer")
for (i in 1:length(dsr_taxa_hits_spp)) {
  dsr_indicies[[i]] <- grep(dsr_taxa_hits_spp[i], rownames(metaphlan2_RA))
}
dsr_indicies <- unlist(dsr_indicies)
dsr_indicies <- unique(dsr_indicies) # sanity check, no repeat rows

# subsetting rel. abund. of cysteine deg bacteria
dsr_m2_RA_controls <- metaphlan2_RA_controls[dsr_indicies, ]
dim(dsr_m2_RA_controls)

# removing rows with __t in the taxa name... leading to double counting of RA
dsr_m2_RA_controls <- dsr_m2_RA_controls[-grep("t__", rownames(dsr_m2_RA_controls)), ]
dim(dsr_m2_RA_controls)

# computing colSums and calculating XX% prevalance of cys. deg. bac.
dsr_m2_cs_controls <- colSums(dsr_m2_RA_controls)
dsr_presnece_prop <- sum(dsr_m2_cs_controls > 0) / length(dsr_m2_cs_controls)
print(paste0("Propotion of samples containing at least one SRB: ", dsr_presnece_prop))
```


# Comparing abundance of SRB and Cys/Met Deg. Bacteria

Below is code for plotting the relative abundances of the various types of h2s producing bacteria from both the `kraken` RAs and the `metaphlan` RAs. At the moment (23 Nov 2020) it appears that we are likely going to go with the kraken values since a kraken DB has already been built for the UHGG representative genomes and it is impossible for us to build a metaphlan database for these genomes. 

## Kraken2 control RAs

```{r}
# prepping data for plotting
df <- data.frame(cysteine = cys_k2_cs_controls, # NOTE: these things are what are imortant for plots
                 methionine = met_k2_cs_controls,
                 cys_met = cys_met_k2_cs_controls,
                 dsr = dsr_k2_cs_controls
                 )

# plotting abundances of sulf(ai)te red bac and cys deg bac
cys_met_dsr_healthy_RA <- ggplot(melt(df), aes(x = factor(variable), y = value, fill = factor(variable))) +
  geom_violin() +
  # geom_boxplot(outlier.alpha = 0.12) + #geom_jitter(aes(alpha = 0.001)) +
  stat_summary(fun = median, geom = "point", shape = 1, size = 4) +
  stat_summary(fun = mean, geom = "point", shape = 4, size = 4) +
  scale_shape_manual("", values=c("mean" = "X")) +
  scale_fill_manual(values = c("#FF0000", "#0000FF", "#7F0080", "#FD9409")) +
  scale_y_continuous(limits = c(0.001, 100),
                     breaks = c(0.01, 0.1, 1, 10, 100),  
                     trans = "log2", 
                     labels = label_comma(accuracy = 0.01)) +
  ggtitle("Relative microbial abundances across H2S production pathways") +
  ylab("relative microbial abundances, log2(%)") + labs(fill = "Function") + 
  theme_bw() +
  theme(axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 15),
        plot.title = element_text(size = 15),
        legend.position = "none") 
cys_met_dsr_healthy_RA
ggsave("../figures/figure2/k2_RAs_healthycontrols.svg", plot = cys_met_dsr_healthy_RA, width = 7, height = 7)
ggsave("../figures/figure2/k2_RAs_healthycontrols.png", plot = cys_met_dsr_healthy_RA, width = 7, height = 7)
```

## MetaPhlAn2 control RAs

```{r}
## TODO: modify to work with metaphlan2 results (currently same as the kraken code)
# prepping data for plotting
df <- data.frame(cysteine = cys_m2_cs_controls, # NOTE: these things are what are imortant for plots
                 methionine = met_m2_cs_controls,
                 cys_met = cys_met_m2_cs_controls,
                 dsr = dsr_m2_cs_controls
                 )

# plotting abundances of sulf(ai)te red bac and cys deg bac
cys_met_dsr_healthy_RA <- ggplot(melt(df), aes(x = factor(variable), y = value, fill = factor(variable))) +
  geom_violin() + 
  # geom_boxplot(outlier.alpha = 0.12) + geom_jitter(aes(alpha = 0.001)) +
  stat_summary(fun = median, geom = "point", shape = 1, size = 4) +
  stat_summary(fun = mean, geom = "point", shape = 4, size = 4) +
  scale_shape_manual("", values=c("mean" = "X")) +
  scale_fill_manual(values = c("#FF0000", "#0000FF", "#7F0080", "#FD9409")) +
  scale_y_continuous(breaks = c(0.01, 0.1, 1, 10, 100),  trans = "log2", labels = label_comma(accuracy = 0.01)) +
  ggtitle("Relative microbial abundances across H2S production pathways") +
  ylab("Relative microbial abundances; log2(%)") + labs(fill = "Function") + 
  theme_bw() +
  theme(axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 15),
        plot.title = element_text(size = 15),
        legend.position = "none") 
cys_met_dsr_healthy_RA
ggsave("../figures/figure2/m2_RAs_healthycontrols.svg", plot = cys_met_dsr_healthy_RA, width = 7, height = 7)
```

## Kraken2 risk RAs 

This next figure will be the second major results figure of the paper and it will compare relative abundances of the 4 differnt classes of h2s producing bacteria across various at risk populations (IBD, CRC and (maybe) patients being treated with AR... this part doesnt matter as much as getting the plot started). 

data processing
```{r}
## calculating full column sums
### cMD
cys_k2_RA <- kraken2_RA[cys_taxa, ]
cys_k2_cs <- colSums(cys_k2_RA)
met_k2_RA <- kraken2_RA[met_taxa, ]
met_k2_cs <- colSums(met_k2_RA)
cys_met_k2_RA <- kraken2_RA[cys_met_taxa, ]
cys_met_k2_cs <- colSums(cys_met_k2_RA)
dsrAB_k2_RA <- kraken2_RA[dsrAB_taxa, ]
dsrAB_k2_cs <- colSums(dsrAB_k2_RA)
### hmp2
cys_k2_hmp2_RA <- kraken2_hmp2_RA[cys_taxa, ]
cys_k2_hmp2_cs <- colSums(cys_k2_hmp2_RA)
met_k2_hmp2_RA <- kraken2_hmp2_RA[met_taxa, ]
met_k2_hmp2_cs <- colSums(met_k2_hmp2_RA)
cys_met_k2_hmp2_RA <- kraken2_hmp2_RA[cys_met_taxa, ]
cys_met_k2_hmp2_cs <- colSums(cys_met_k2_hmp2_RA)
dsrAB_k2_hmp2_RA <- kraken2_hmp2_RA[dsrAB_taxa, ]
dsrAB_k2_hmp2_cs <- colSums(dsrAB_k2_hmp2_RA)

## filtering pData to only include populations for plotting
# kraken2_pData %>%
#   filter(study_condition == c("IBD", "CRC", "control")) -> kraken_risk_pData
kraken2_pData %>%
  filter(study_condition == c("IBD", "CRC")) -> ibd_crc
unique(ibd_crc$dataset_name) -> ibd_crc_studies
kraken_risk_pData <- kraken2_pData[kraken2_pData$dataset_name %in% ibd_crc_studies, ]
# kraken2_hmp2_pData ## already subsetted to risk populations
### cMD
cys_k2_cs_risk <- cys_k2_cs[names(cys_k2_cs) %in% rownames(kraken_risk_pData)]
met_k2_cs_risk <- met_k2_cs[names(met_k2_cs) %in% rownames(kraken_risk_pData)]
cys_met_k2_cs_risk <- cys_met_k2_cs[names(cys_met_k2_cs) %in% rownames(kraken_risk_pData)]
dsrAB_k2_cs_risk <- dsrAB_k2_cs[names(dsrAB_k2_cs) %in% rownames(kraken_risk_pData)]
### HMP2
cys_k2_hmp2_cs_risk <- cys_k2_hmp2_cs[names(cys_k2_hmp2_cs) %in% kraken2_hmp2_pData$External.ID.mod]
tmp <- kraken2_hmp2_pData[kraken2_hmp2_pData$External.ID.mod %in% names(cys_k2_hmp2_cs),]
met_k2_hmp2_cs_risk <- met_k2_hmp2_cs[names(met_k2_hmp2_cs) %in% kraken2_hmp2_pData$External.ID.mod]
cys_met_k2_hmp2_cs_risk <- cys_met_k2_hmp2_cs[names(cys_met_k2_hmp2_cs) %in% kraken2_hmp2_pData$External.ID.mod]
dsrAB_k2_hmp2_cs_risk <- dsrAB_k2_hmp2_cs[names(dsrAB_k2_hmp2_cs) %in% kraken2_hmp2_pData$External.ID.mod]

## making the necessary dataframes
### cMD
hmp2_study_cond <- as.character(kraken_risk_pData$study_condition)
names(hmp2_study_cond) <- rownames(kraken_risk_pData)

cys_k2_df <- as.data.frame(cbind(cys_k2_cs_risk, hmp2_study_cond = hmp2_study_cond[names(cys_k2_cs_risk)]))
colnames(cys_k2_df) <- c("RA", "population")
cys_k2_df$RA <- as.numeric(cys_k2_df$RA)
cys_k2_df$func <- rep("cysteine", length(cys_k2_cs_risk))
cys_k2_df$population <- factor(cys_k2_df$population, levels = c("control", "IBD", "adenoma", "CRC"))

met_k2_df <- as.data.frame(cbind(met_k2_cs_risk, hmp2_study_cond = hmp2_study_cond[names(met_k2_cs_risk)]))
colnames(met_k2_df) <- c("RA", "population")
met_k2_df$RA <- as.numeric(met_k2_df$RA)
met_k2_df$func <- rep("methionine", length(met_k2_cs_risk))
met_k2_df$population <- factor(met_k2_df$population, levels = c("control", "IBD", "adenoma", "CRC"))

cys_met_k2_df <- as.data.frame(cbind(cys_met_k2_cs_risk, hmp2_study_cond = hmp2_study_cond[names(cys_met_k2_cs_risk)]))
colnames(cys_met_k2_df) <- c("RA", "population")
cys_met_k2_df$RA <- as.numeric(cys_met_k2_df$RA)
cys_met_k2_df$func <- rep("cys_met", length(cys_met_k2_cs_risk))
cys_met_k2_df$population <- factor(cys_met_k2_df$population, levels = c("control", "IBD", "adenoma", "CRC"))

dsrAB_k2_df <- as.data.frame(cbind(dsrAB_k2_cs_risk, hmp2_study_cond = hmp2_study_cond[names(dsrAB_k2_cs_risk)]))
colnames(dsrAB_k2_df) <- c("RA", "population")
dsrAB_k2_df$RA <- as.numeric(dsrAB_k2_df$RA)
dsrAB_k2_df$func <- rep("dsrAB", length(dsrAB_k2_cs_risk))
dsrAB_k2_df$population <- factor(dsrAB_k2_df$population, levels = c("control", "IBD", "adenoma", "CRC"))

# cys_k2_df <- data.frame(func = rep("cysteine", length(cys_k2_cs_risk)), 
#                  RA = cys_k2_cs_risk,
#                  population = kraken_risk_pData$study_condition)
# met_k2_df <- data.frame(func = rep("methionine", length(met_k2_cs_risk)), 
#                  RA = met_k2_cs_risk,
#                  population = kraken_risk_pData$study_condition)
# cys_met_k2_df <- data.frame(func = rep("cys_and_met", length(cys_met_k2_cs_risk)), 
#                  RA = cys_met_k2_cs_risk,
#                  population = kraken_risk_pData$study_condition)
# dsrAB_k2_df <- data.frame(func = rep("dsr", length(dsrAB_k2_cs_risk)), 
#                  RA = dsrAB_k2_cs_risk,
#                  population = kraken_risk_pData$study_condition)

### HMP2
subset_k2_pData <- select(kraken2_hmp2_pData, External.ID, diagnosis)
k2_risk_pData <- subset_k2_pData$diagnosis
names(k2_risk_pData) <- subset_k2_pData$External.ID

cys_k2_hmp2_df <- as.data.frame(cbind(cys_k2_hmp2_cs_risk, k2_risk_pData = k2_risk_pData[names(cys_k2_hmp2_cs_risk)]))
colnames(cys_k2_hmp2_df) <- c("RA", "diagnosis")
cys_k2_hmp2_df$RA <- as.numeric(cys_k2_hmp2_df$RA)
cys_k2_hmp2_df$func <- rep("cysteine", length(cys_k2_hmp2_cs_risk))
cys_k2_hmp2_df$diagnosis <- factor(cys_k2_hmp2_df$diagnosis, levels = c("nonIBD", "UC", "CD"))
# colnames(cys_k2_hmp2_df) <- c("RA", "diagnosis", "func")

met_k2_hmp2_df <- as.data.frame(cbind(met_k2_hmp2_cs_risk, k2_risk_pData = k2_risk_pData[names(met_k2_hmp2_cs_risk)]))
colnames(met_k2_hmp2_df) <- c("RA", "diagnosis")
met_k2_hmp2_df$RA <- as.numeric(met_k2_hmp2_df$RA)
met_k2_hmp2_df$func <- rep("methionine", length(met_k2_hmp2_cs_risk))
met_k2_hmp2_df$diagnosis <- factor(met_k2_hmp2_df$diagnosis, levels = c("nonIBD", "UC", "CD"))
# colnames(met_k2_hmp2_df) <- c("RA", "diagnosis", "func")

cys_met_k2_hmp2_df <- as.data.frame(cbind(cys_met_k2_hmp2_cs_risk, k2_risk_pData = k2_risk_pData[names(cys_met_k2_hmp2_cs_risk)]))
colnames(cys_met_k2_hmp2_df) <- c("RA", "diagnosis")
cys_met_k2_hmp2_df$RA <- as.numeric(cys_met_k2_hmp2_df$RA)
cys_met_k2_hmp2_df$func <- rep("cys_met", length(cys_met_k2_hmp2_cs_risk))
cys_met_k2_hmp2_df$diagnosis <- factor(cys_met_k2_hmp2_df$diagnosis, levels = c("nonIBD", "UC", "CD"))
# colnames(cys_met_k2_hmp2_df) <- c("RA", "diagnosis", "func")

dsrAB_k2_hmp2_df <- as.data.frame(cbind(dsrAB_k2_hmp2_cs_risk, k2_risk_pData = k2_risk_pData[names(dsrAB_k2_hmp2_cs_risk)]))
colnames(dsrAB_k2_hmp2_df) <- c("RA", "diagnosis")
dsrAB_k2_hmp2_df$RA <- as.numeric(dsrAB_k2_hmp2_df$RA)
dsrAB_k2_hmp2_df$func <- rep("dsrAB", length(dsrAB_k2_hmp2_cs_risk))
dsrAB_k2_hmp2_df$diagnosis <- factor(dsrAB_k2_hmp2_df$diagnosis, levels = c("nonIBD", "UC", "CD"))
# colnames(dsrAB_k2_hmp2_df) <- c("RA", "diagnosis", "func")

# cys_k2_hmp2_df <- data.frame(func = rep("cysteine", length(cys_k2_hmp2_cs_risk)), 
#                  RA = cys_k2_hmp2_cs_risk,
#                  population = kraken2_hmp2_pData$diagnosis)
# met_k2_hmp2_df <- data.frame(func = rep("methionine", length(met_k2_hmp2_cs_risk)), 
#                  RA = met_k2_hmp2_cs_risk,
#                  population = kraken2_hmp2_pData$diagnosis)
# cys_met_k2_hmp2_df <- data.frame(func = rep("cys_and_met", length(cys_met_k2_hmp2_cs_risk)), 
#                  RA = cys_met_k2_hmp2_cs_risk,
#                  population = kraken2_hmp2_pData$diagnosis)
# dsrAB_k2_hmp2_df <- data.frame(func = rep("dsr", length(dsrAB_k2_hmp2_cs_risk)), 
#                  RA = dsrAB_k2_hmp2_cs_risk,
#                  population = kraken2_hmp2_pData$diagnosis)
```

plotting data
cMD
```{r}
clrs <- c("#999999", "#E69F00", "#56B4E9", "#0072B2")
cys_k2_boxplot <- ggplot(cys_k2_df) + 
  geom_boxplot(aes(x = func, y = RA, fill = population)) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0.001, 100),
                     breaks = c(0.01, 0.1, 1, 10, 100),
                     trans = "log2",
                     minor_breaks = NULL) +
  scale_fill_manual(values = clrs) +
  ylab("relative bacterial abundance, log2[%]") + xlab("")
# cys_k2_boxplot
met_k2_boxplot <- ggplot(met_k2_df) + 
  geom_boxplot(aes(x = func, y = RA, fill = population)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  scale_y_continuous(limits = c(0.001, 100),
                     breaks = c(0.01, 0.1, 1, 10, 100),
                     trans = "log2", 
                     minor_breaks = NULL) +
  scale_fill_manual(values = clrs) +
  xlab("")
# met_k2_boxplot
cys_met_k2_boxplot <- ggplot(cys_met_k2_df) + 
  geom_boxplot(aes(x = func, y = RA, fill = population)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  scale_y_continuous(limits = c(0.001, 100),
                     breaks = c(0.01, 0.1, 1, 10, 100),
                     trans = "log2", 
                     minor_breaks = NULL) +
  scale_fill_manual(values = clrs) +
  xlab("")
# cys_met_k2_boxplot
dsrAB_k2_boxplot <- ggplot(dsrAB_k2_df) + 
  geom_boxplot(aes(x = func, y = RA, fill = population)) +
  theme_bw() +
  theme(axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  scale_y_continuous(limits = c(0.001, 100),
                     breaks = c(0.01, 0.1, 1, 10, 100),
                     trans = "log2", 
                     minor_breaks = NULL) +
  scale_fill_manual(values = clrs) +

  xlab("")
# dsrAB_k2_boxplot

# =========================================================================== #

## ARRANGING PLOTS 
cys_k2_boxplot | met_k2_boxplot | cys_met_k2_boxplot | dsrAB_k2_boxplot
ggsave("../figures/figure3/k2_RA_risk_populations.svg", height = 4, width = 7)
ggsave("../figures/figure3/k2_RA_risk_populations.png", height = 4, width = 7)
```

hmp2
```{r}
# p <- ggplot(cys_k2_hmp2_df) + 
#   geom_boxplot(aes(x = func, y = RA, fill = diagnosis))

clrs <- c("#999999", "#F0E442", "#D55E00")
cys_k2_hmp2_boxplot <- ggplot(cys_k2_hmp2_df) + 
  geom_boxplot(aes(x = func, y = RA, fill = diagnosis)) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(0.001, 100),
                     breaks = c(0.01, 0.1, 1, 10, 100),
                     trans = "log2") +
  scale_fill_manual(values = clrs) +
  ylab("relative bacterial abundance, log2[%]") + xlab("")
# cys_k2_hmp2_boxplot
met_k2_hmp2_boxplot <- ggplot(met_k2_hmp2_df) + 
  geom_boxplot(aes(x = func, y = RA, fill = diagnosis)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  scale_y_continuous(limits = c(0.001, 100),
                     breaks = c(0.01, 0.1, 1, 10, 100),
                     trans = "log2", 
                     minor_breaks = NULL) +
  scale_fill_manual(values = clrs) +
  xlab("")
# met_k2_hmp2_boxplot
cys_met_k2_hmp2_boxplot <- ggplot(cys_met_k2_hmp2_df) + 
  geom_boxplot(aes(x = func, y = RA, fill = diagnosis)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  scale_y_continuous(limits = c(0.001, 100),
                     breaks = c(0.01, 0.1, 1, 10, 100),
                     trans = "log2", 
                     minor_breaks = NULL) +
  scale_fill_manual(values = clrs) +
  xlab("")
# cys_met_k2_hmp2_boxplot
dsrAB_k2_hmp2_boxplot <- ggplot(dsrAB_k2_hmp2_df) + 
  geom_boxplot(aes(x = func, y = RA, fill = diagnosis)) +
  theme_bw() +
  theme(axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  scale_y_continuous(limits = c(0.001, 100),
                     breaks = c(0.01, 0.1, 1, 10, 100),
                     trans = "log2", 
                     minor_breaks = NULL) +
  scale_fill_manual(values = clrs) +

  xlab("")
# dsrAB_k2_hmp2_boxplot

# =========================================================================== #

## ARRANGING PLOTS 
cys_k2_hmp2_boxplot | met_k2_hmp2_boxplot | cys_met_k2_hmp2_boxplot | dsrAB_k2_hmp2_boxplot
ggsave("../figures/figure3/k2_hmp2_RA_risk_populations.svg", height = 4, width = 7)
ggsave("../figures/figure3/k2_hmp2_RA_risk_populations.png", height = 4, width = 7)
```

## MetaPhlAn2 risk RAs

data processing
```{r}
## setting function indicies
cys_m2_RA <- metaphlan2_RA[cys_indicies, ]
met_m2_RA <- metaphlan2_RA[met_indicies, ]
cys_met_m2_RA <- metaphlan2_RA[both_indicies, ]
dsr_m2_RA <- metaphlan2_RA[dsr_indicies, ]

## removing rows with __t in the taxa name... leading to double counting of RA
cys_m2_RA <- cys_m2_RA[-grep("t__", rownames(cys_m2_RA)), ]
cys_m2_cs <- colSums(cys_m2_RA)
met_m2_RA <- met_m2_RA[-grep("t__", rownames(met_m2_RA)), ]
met_m2_cs <- colSums(met_m2_RA)
cys_met_m2_RA <- cys_met_m2_RA[-grep("t__", rownames(cys_met_m2_RA)), ]
cys_met_m2_cs <- colSums(cys_met_m2_RA)
dsr_m2_RA <- dsr_m2_RA[-grep("t__", rownames(dsr_m2_RA)), ]
dsr_m2_cs <- colSums(dsr_m2_RA)

## filtering pData to only include populations for plotting
metaphlan2_pData %>%
  filter(study_condition == c("IBD", "CRC", "control")) -> risk_pData
cys_risk_cs <- cys_m2_cs[names(cys_m2_cs) %in% rownames(risk_pData)]
met_risk_cs <- met_m2_cs[names(met_m2_cs) %in% rownames(risk_pData)]
cys_met_risk_cs <- cys_met_m2_cs[names(cys_met_m2_cs) %in% rownames(risk_pData)]
dsr_risk_cs <- dsr_m2_cs[names(dsr_m2_cs) %in% rownames(risk_pData)]
# metaphlan2_pData %>%
#   filter(study_condition == c("IBD", "CRC")) -> ibd_crc
# unique(ibd_crc$studyID) -> ibd_crc_studies
# ibd_crc_pData <- metaphlan2_pData[metaphlan2_pData$studyID %in% ibd_crc_studies, ]

## making the necessary dataframe
cys_df <- data.frame(func = rep("cysteine", length(cys_risk_cs)), 
                 RA = cys_risk_cs,
                 population = risk_pData$study_condition)
met_df <- data.frame(func = rep("methionine", length(met_risk_cs)), 
                 RA = met_risk_cs,
                 population = risk_pData$study_condition)
cys_met_df <- data.frame(func = rep("cys_and_met", length(cys_met_risk_cs)), 
                 RA = cys_met_risk_cs,
                 population = risk_pData$study_condition)
dsr_df <- data.frame(func = rep("dsr", length(dsr_risk_cs)), 
                 RA = dsr_risk_cs,
                 population = risk_pData$study_condition)
```

plotting data
```{r}
clrs <- c("grey", "gold2", "green3")
cys_boxplot <- ggplot(cys_df) + 
  geom_boxplot(aes(x = func, y = RA, fill = population)) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_y_continuous(
                     breaks = c(1, 10, 100),
                     trans = "log2") +
  scale_fill_manual(values = clrs) +
  ylab("Relative Abundance (log2(%))") + xlab("")
# cys_boxplot
met_boxplot <- ggplot(met_df) + 
  geom_boxplot(aes(x = func, y = RA, fill = population)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  scale_y_continuous(limits = c(0.0001, 100),
                     breaks = c(1, 10, 100),
                     trans = "log2", 
                     minor_breaks = NULL) +
  scale_fill_manual(values = clrs) +
  xlab("")
# met_boxplot
cys_met_boxplot <- ggplot(cys_met_df) + 
  geom_boxplot(aes(x = func, y = RA, fill = population)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  scale_y_continuous(limits = c(0.001, 100),
                     breaks = c(1, 10, 100), 
                     trans = "log2", 
                     minor_breaks = NULL) +
  scale_fill_manual(values = clrs) +
  xlab("")
# cys_met_boxplot
dsr_boxplot <- ggplot(dsr_df) + 
  geom_boxplot(aes(x = func, y = RA, fill = population)) +
  theme_bw() +
  theme(axis.title.y = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) +
  scale_y_continuous(limits = c(0.001, 100),
                     breaks = c(1, 10, 100), 
                     trans = "log2", 
                     minor_breaks = NULL) +
  scale_fill_manual(values = clrs) +

  xlab("")
# dsr_boxplot

# =========================================================================== #

## ARRANGING PLOTS 
cys_boxplot | met_boxplot | cys_met_boxplot | dsr_boxplot
ggsave("../figures/figure3/m2_RA_across_risk_populations.svg", height = 4, width = 7)
ggsave("../figures/figure3/m2_RA_across_risk_populations.png", height = 4, width = 7)
```

# Basement

contains depreciated code, but things that I want to keep around as reference to previous results and plots.

## Prevalence of sulfite reducing bacteria in healthy controls

```{r}
# # getting B. wadsworthia prevalence
# b_wad <- control_RA[grep("s__Bilophila_wadsworthia$", rownames( control_RA)), ]
# b_wad_summary <- summary(b_wad)
# b_wad_sd <- sd(b_wad)
# b_wad_prev <- sum(b_wad > 0) / length(b_wad)
```

## Prevalence of sulfate reducing bacteria

```{r}
# ## getting Desulfovibrio spp: piger, desulfuricans, gigas
# d_piger <- control_RA[grep("s__Desulfovibrio_piger$", rownames(control_RA)), ]
# d_piger_summary <- summary(d_piger)
# d_piger_sd <- sd(d_piger)
# d_piger_prev <- sum(d_piger > 0) / length(d_piger)
# 
# ## desulfricans
# d_desulf <- control_RA[grep("s__Desulfovibrio_desulfuricans$", rownames(control_RA)), ]
# d_desulf_summary <- summary(d_desulf)
# d_desulf_sd <- sd(d_desulf)
# d_desulf_prev <- sum(d_desulf > 0) / length(d_desulf)
# 
# ## gigas
# d_gigas <- control_RA[grep("s__Desulfovibrio_gigas$", rownames(control_RA)), ]
# d_gigas_summary <- summary(d_gigas)
# d_gigas_sd <- sd(d_gigas)
# d_gigas_prev <- sum(d_gigas > 0) / length(d_gigas)
# 
# ## presence of one or more of above 4 species
# sulf_red_RA <- data.frame(b_wad = b_wad, 
#                        d_piger = d_piger,
#                        d_desulf = d_desulf,
#                        d_gigas = d_gigas)
# sulf_red_RA_rs <- rowSums(sulf_red_RA)
# sulf_red_RA_prev <- sum(sulf_red_RA_rs > 0) / length(sulf_red_RA_rs)
```

## Comparison of cysteine degrading and methionine degrading bacteria

```{r}
# ## load in pie chart file
# cys_met_overlap <- read.table("../results/from-GutFunFind/H2S_production.pichart.txt", skip = 7)
# rownames(cys_met_overlap) <- cys_met_overlap$V1
# cys_met_overlap %>%
#   select(V4, V5, V6, V7) -> cys_met_overlap
# colnames(cys_met_overlap) <- c("both", "absent", "methionine", "cysteine")
# set1 <- cys_met_overlap %>% filter(methionine > 0) %>% rownames()
# set2 <- cys_met_overlap %>% filter(cysteine > 0) %>% rownames()
# # set3 <- tmp %>% filter(both > 0) %>% rownames()
# 
# ## venn diagram showing overlap between cys and met degrading functionality
# venn.diagram(
#   x = list(set1, set2),
#   category.names = c("methionine" , "cysteine"),
#   filename = "../figures/cys_met_overlap.png",
#   output = TRUE,
#           imagetype="png" ,
#           height = 480,
#           width = 480,
#           resolution = 300,
#           lwd = 1,
#           col = c("blue", "red"),
#           fill = c(alpha("blue",0.3), alpha('red',0.3)),
#           cex = 0.5,
#           fontfamily = "sans",
#           cat.cex = 0.3,
#           cat.default.pos = "outer",
#           cat.pos = c(-27, 27),
#           cat.dist = c(0.055, 0.055),
#           cat.fontfamily = "sans",
#           cat.col = c("blue", "red"),
# )
```

## Processing taxa_hits from GutFunFind to match with metaphlan2 results

The below code takes an initial 493 taxa hits from `GutFunFind` and filters out names without species level resolition. A number of other modifications were also made to make the taxonomic name searchable against the identifiers present in `curatedMetagenomicData`. The variable `taxa_hits_spp` contains a final list of 254 Genus_species names with cysteine degrading potential.

```{r}
# cys_met_taxa_hits <- as.vector(t(cys_met_taxa_hits_raw), mode = "character") # df -> vector
# cys_met_taxa_hits <- strsplit(cys_met_taxa_hits, ";")
# 
# ## extracting species names from taxa hits
# cys_met_taxa_hits_spp <- vector(mode = "character", length = length(cys_met_taxa_hits)) # create empty caracter vector
# for (i in 1:length(cys_met_taxa_hits)) {
#   current_spp <- gsub(pattern = "s__", replacement = "", cys_met_taxa_hits[[i]][7])
#   cys_met_taxa_hits_spp[i] <- current_spp
# }
# 
# ## subetting feature table to non-absent representative genomes (contained in dsr_taxa_hits_raw)
# filter(cys_met_overlap, absent == 0) -> cys_met_overlap_present
# 
# ## removing empty species names
# cys_met_overlap_present <- cys_met_overlap_present[cys_met_taxa_hits_spp != "", ] # simultaneously subsetting the feature table
# cys_met_taxa_hits_spp <- cys_met_taxa_hits_spp[cys_met_taxa_hits_spp != ""]
# dim(cys_met_overlap_present)
# length(cys_met_taxa_hits_spp)
# 
# ## removing Co-Abundance gene Groups (CAGs), GCAs (?), etc.
# if (length(grep("CAG", cys_met_taxa_hits_spp)) == 0) {
#   # do nothing
# } else if (length(grep("CAG", cys_met_taxa_hits_spp)) > 0) {
#   cys_met_overlap_present <- cys_met_overlap_present[-grep("CAG", cys_met_taxa_hits_spp), ]
#   cys_met_taxa_hits_spp <- cys_met_taxa_hits_spp[-grep("CAG", cys_met_taxa_hits_spp)]
# }
# dim(cys_met_overlap_present)
# length(cys_met_taxa_hits_spp)
# if (length(grep("GCA", cys_met_taxa_hits_spp)) == 0) {
#   # do nothing
# } else if (length(grep("GCA", cys_met_taxa_hits_spp)) > 0) {
#   cys_met_overlap_present <- cys_met_overlap_present[-grep("GCA", cys_met_taxa_hits_spp), ]
#   cys_met_taxa_hits_spp <- cys_met_taxa_hits_spp[-grep("GCA", cys_met_taxa_hits_spp)]
# }
# dim(cys_met_overlap_present)
# length(cys_met_taxa_hits_spp)
# if (length(grep("sp[0-9]*", cys_met_taxa_hits_spp)) == 0) {
#   # do nothing
# } else if (length(grep("sp[0-9]*", cys_met_taxa_hits_spp)) > 0) {
#   cys_met_overlap_present <- cys_met_overlap_present[-grep("sp[0-9]*", cys_met_taxa_hits_spp), ]
#   cys_met_taxa_hits_spp <- cys_met_taxa_hits_spp[-grep("sp[0-9]*", cys_met_taxa_hits_spp)]
# }
# dim(cys_met_overlap_present)
# length(cys_met_taxa_hits_spp)
# 
# ## subetting based on function hits (complicated for cys met)
# 
# ### cys degrading
# cys_taxa_hits_spp <- cys_met_taxa_hits_spp[cys_met_overlap_present$cysteine == 1]
# cys_overlap_present <- cys_met_overlap_present[cys_met_overlap_present$cysteine == 1, ]
# dim(cys_overlap_present)
# length(cys_taxa_hits_spp)
# #### removing oddities from names like "_A", "_X", etc.
# for (i in 1:length(cys_taxa_hits_spp)) {
#   current_name <- gsub("_.", "", cys_taxa_hits_spp[i])
#   current_name <- gsub(" ", "_", current_name)
#   cys_taxa_hits_spp[i] <- current_name
# }
# #### removing duplicates of species names
# cys_taxa_hits_spp <- unique(cys_taxa_hits_spp)
# length(cys_taxa_hits_spp)
# 
# ### met degrading
# met_taxa_hits_spp <- cys_met_taxa_hits_spp[cys_met_overlap_present$methionine == 1]
# met_overlap_present <- cys_met_overlap_present[cys_met_overlap_present$methionine == 1, ]
# dim(met_overlap_present)
# length(met_taxa_hits_spp)
# #### removing oddities from names like "_A", "_X", etc.
# for (i in 1:length(met_taxa_hits_spp)) {
#   current_name <- gsub("_.", "", met_taxa_hits_spp[i])
#   current_name <- gsub(" ", "_", current_name)
#   met_taxa_hits_spp[i] <- current_name
# }
# #### removing duplicates of species names
# met_taxa_hits_spp <- unique(met_taxa_hits_spp)
# length(met_taxa_hits_spp)
# 
# ### both degrading
# both_taxa_hits_spp <- cys_met_taxa_hits_spp[(cys_met_overlap_present$methionine == 1 & cys_met_overlap_present$cysteine == 1)]
# both_overlap_present <- cys_met_overlap_present[(cys_met_overlap_present$methionine == 1 & cys_met_overlap_present$cysteine == 1), ]
# dim(both_overlap_present)
# length(both_taxa_hits_spp)
# #### removing oddities from names like "_A", "_X", etc.
# for (i in 1:length(both_taxa_hits_spp)) {
#   current_name <- gsub("_.", "", both_taxa_hits_spp[i])
#   current_name <- gsub(" ", "_", current_name)
#   both_taxa_hits_spp[i] <- current_name
# }
# #### removing duplicates of species names
# both_taxa_hits_spp <- unique(both_taxa_hits_spp)
# length(both_taxa_hits_spp)
```

```{r}
# dsr_taxa_hits <- as.vector(t(dsr_taxa_hits_raw), mode = "character") # df -> vector
# dsr_taxa_hits <- strsplit(dsr_taxa_hits, ";")
# tmp <- unlist(dsr_taxa_hits)
# ## extracting species names from taxa hits
# dsr_taxa_hits_spp <- vector(mode = "character", length = length(dsr_taxa_hits)) # create empty caracter vector
# for (i in 1:length(dsr_taxa_hits)) {
#   current_spp <- gsub(pattern = "s__", replacement = "", dsr_taxa_hits[[i]][7])
#   dsr_taxa_hits_spp[i] <- current_spp
# }
# 
# ## subetting feature table to non-absent representative genomes (contained in dsr_taxa_hits_raw)
# filter(dsr_overlap, absent == 0) -> dsr_overlap_present
# 
# ### for use in debugging gutfunfind results for diss. sulf.
# # tmp_taxa <- dsr_taxa_hits[dsr_overlap_present$dsrAB == 1] ## NOTE: the version of `dsr_taxa_hits` required for this step
# #                                                           ## is the version outputted immediately after the _raw line
# # tmp_func <- filter(dsr_overlap_present, dsrAB == 1)
# # tmp_prob_genID <- tmp_func$genome[c(2,5,6,7,11,17,18,19,20,26)]
# # tmp_problem_taxa <- tmp_taxa[c(2,5,6,7,11,17,18,19,20,26)]
# # tmp_accep_genID <- tmp_func$genome[c(1,3,4,8,9,10,12,13,14,15,16,21,22,23,24,25,27)]
# # tmp_accepted_taxa <- tmp_taxa[c(1,3,4,8,9,10,12,13,14,15,16,21,22,23,24,25,27)]
# 
# ## removing empty species names
# dsr_overlap_present <- dsr_overlap_present[dsr_taxa_hits_spp != "", ] # simultaneously subsetting the feature table
# dsr_taxa_hits_spp <- dsr_taxa_hits_spp[dsr_taxa_hits_spp != ""]
# 
# ## removing Co-Abundance gene Groups (CAGs), GCAs (?), etc.
# if (length(grep("CAG", dsr_taxa_hits_spp)) == 0) {
#   # do nothing
# } else if (length(grep("CAG", dsr_taxa_hits_spp)) > 0) {
#   dsr_overlap_present <- dsr_overlap_present[-grep("CAG", dsr_taxa_hits_spp), ]
#   dsr_taxa_hits_spp <- dsr_taxa_hits_spp[-grep("CAG", dsr_taxa_hits_spp)]
# }
# dim(dsr_overlap_present)
# length(dsr_taxa_hits_spp)
# if (length(grep("GCA", dsr_taxa_hits_spp)) == 0) {
#   # do nothing
# } else if (length(grep("GCA", dsr_taxa_hits_spp)) > 0) {
#   dsr_overlap_present <- dsr_overlap_present[-grep("GCA", dsr_taxa_hits_spp), ]
#   dsr_taxa_hits_spp <- dsr_taxa_hits_spp[-grep("GCA", dsr_taxa_hits_spp)]
# }
# dim(dsr_overlap_present)
# length(dsr_taxa_hits_spp)
# if (length(grep("sp[0-9]*", dsr_taxa_hits_spp)) == 0) {
#   # do nothing
# } else if (length(grep("sp[0-9]*", dsr_taxa_hits_spp)) > 0) {
#   dsr_overlap_present <- dsr_overlap_present[-grep("sp[0-9]*", dsr_taxa_hits_spp), ]
#   dsr_taxa_hits_spp <- dsr_taxa_hits_spp[-grep("sp[0-9]*", dsr_taxa_hits_spp)]
# }
# dim(dsr_overlap_present)
# length(dsr_taxa_hits_spp)
# 
# ## subsetting based on function hits (for example, dsrAB must be present)
# dsr_taxa_hits_spp <- dsr_taxa_hits_spp[dsr_overlap_present$dsrAB == 1]
# dsr_overlap_present <- dsr_overlap_present[dsr_overlap_present$dsrAB == 1, ]
# dim(dsr_overlap_present)
# length(dsr_taxa_hits_spp)
# 
# ## removing oddities from names like "_A", "_X", etc.
# for (i in 1:length(dsr_taxa_hits_spp)) {
#   current_name <- gsub("_.", "", dsr_taxa_hits_spp[i])
#   current_name <- gsub(" ", "_", current_name)
#   dsr_taxa_hits_spp[i] <- current_name
# }
# length(dsr_taxa_hits_spp)
# 
# ## removing duplicates of species names
# dsr_taxa_hits_spp <- unique(dsr_taxa_hits_spp)
# length(dsr_taxa_hits_spp)
```

```{r}
# indicies <- list(mode = "integer")
# for (i in 1:length(dsr_taxa_hits_spp)) {
#   indicies[[i]] <- grep(dsr_taxa_hits_spp[i], rownames(control_RA))
# }
# indicies <- unlist(indicies)
# indicies <- unique(indicies) # sanity check, no repeat rows
# 
# # subsetting rel. abund. of dsr bacteria
# dsr_RA <- control_RA[indicies, ]
# dim(dsr_RA)
# 
# # removing rows with __t in the taxa name... leading to double counting of RA
# dsr_RA <- dsr_RA[-grep("t__", rownames(dsr_RA)), ]
# dim(dsr_RA)
# 
# # computing colSums and calculating XX% prevalance of cys. deg. bac.
# dsr_cs <- colSums(dsr_RA)
# sum(dsr_cs > 0) / length(dsr_cs)
```
